Tool-driven Revolutions in Archaeological Science

### 2019-08-12 analysis/paper/jcaa-paper.Rmd
resubmit to JCAA with response to editor and peer reviewers

min_y <- 2008
max_y <- 2021
label_text_size = 2

@@

all_archaeology_citing_R_over_time <-
  ggplot(archy_cran_cites_all_areas_top_journals_by_year_to_plot,
         aes(PY,
             prop,
             colour = SO)) +
  scale_x_continuous(breaks = min_y:2018,
                     limits = c(min_y, max_y)) +
  scale_y_continuous(limits = c(0, 0.2)) +
  geom_line(size = 2) +
  geom_point() +
  geom_text_repel(
    data = archy_cran_cites_all_areas_top_journals_by_year_to_plot %>%
      group_by(SO) %>%
      slice(which.max(PY)) %>% 
      mutate(SO_n = str_wrap(SO_n, 60)),
    aes(label = SO_n),
    hjust = -1,
    direction = "y",
    size = label_text_size,
    nudge_x = 2) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.title.x = element_text(hjust = 0.4),
        legend.position="none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5)) +
  xlab("Publication year") +
  ylab("Proportion of articles in that journal") +
  lemon::coord_capped_cart(bottom='right')

@@

To compare with ecology we conducted a similar bibliometric analysis for archaeology journals indexed by the Web of Science service. We used the 'Cited Reference Search' to find articles citing R in the same way as above, and then refined the results to keep only those articles published during 2008-2018 that are included in the Web of Science category 'Archaeology'. This resulted in `r prettyNum(how_many_articles_in_top_archy_journals, ",")` articles, of which `r total_number_archaeology_articles_citing_r` cite R.  Figure \@ref(fig:archaeology-journals-with-subplot) shows the temporal trend of citations of R in archaeological articles. There are three interesting details revealed by this figure.

```{r archaeology-journals-with-subplot, fig.cap=str_glue('Proportion of Archaeology articles per year citing R (a total of {total_number_archaeology_articles_citing_r} out of {prettyNum(how_many_articles_in_top_archy_journals, ",")} articles in our sample for 2008-2018). Sub-plot shows articles published in the _Journal of Archaeological Science_ during 2008-2017. Data from Web of Science')}
# combine main plot and subplot
all_archaeology_citing_R_over_time_and_subplot <-
all_archaeology_citing_R_over_time  +
  annotation_custom(grob = ggplotGrob(jas_cites_r),
                          xmin = 2008, xmax = 2015,
                          ymin = 0.08, ymax = 0.20)
ggsave(plot = all_archaeology_citing_R_over_time_and_subplot,
       filename = "all_archaeology_citing_R_over_time_and_subplot.png",
       path = str_glue('{here::here()}/analysis/figures'),
       height = 9,
       width = 12)
knitr::include_graphics(str_glue('{here::here()}/analysis/figures/all_archaeology_citing_R_over_time_and_subplot.png'))
```

@@

As with data (see Fitzpatrick in this volume), preparing code to make it publicly available takes time to ensure that it is fit for others to read and use. Our casual observations of the ways in which archaeologists are using and sharing code show a high degree of variability in code style and organisation, indicating that currently most archaeologists are independently solving problems of how to write and share code through trial and error. We can use concepts from cultural evolution to understand this situation: learning by trial and error is known as guided variation in a cultural evolutionary framework. The dominance of guided variation suggests that using code is not yet widespread enough among archaeologists to propagate due to frequency-dependent biases. Frequency-dependent biases occur when people copy the most abundant variant, in this case the most commonly used tool for analysing data, in the population [cf. @boyd1988culture]. It also suggests that the professional benefits (i.e. the citation effects we describe above) are not yet widely known for code use to propagate due to content biases. Content biases result when some aspect of a variant's content, such as positive citation effects, makes it more likely to be adopted. Finally there may not yet enough highly-visible, prestigous researchers using code for this behaviour due to propagate due to model-based biases [@Rendell_et_al_2011; @boyd1988culture; @henrich2007dual]. This bias is based on imitation of highly prestigous, skilled or successful individuals. Considering the high variability of code using cultural evolutionary concepts, we conclude that there is unrealised potential to improve the efficiency of using and code in archaeology, firstly by converging on some widely-agreed upon conventions that will save researcher's time, and secondly by investing effort in communicating the professional benefits of using code.

>>>>

min_y <- 2008
max_y <- 2021
label_text_size = 4

@@

all_archaeology_citing_R_over_time <-
  ggplot(archy_cran_cites_all_areas_top_journals_by_year_to_plot,
         aes(PY,
             prop,
             colour = SO)) +
  scale_x_continuous(breaks = min_y:2018,
                     limits = c(min_y, max_y)) +
  scale_y_continuous(limits = c(0, 0.2)) +
  geom_line(size = 2) +
  geom_point() +
  geom_text_repel(
    data = archy_cran_cites_all_areas_top_journals_by_year_to_plot %>%
      group_by(SO) %>%
      slice(which.max(PY)) %>% 
      ungroup() %>% 
      top_n(5, total_n_articles) %>% 
      mutate(SO_n = str_wrap(SO_n, 30)),
    aes(label = SO_n),
    hjust = -1,
    direction = "y",
    size = label_text_size,
    nudge_x = 5) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        axis.title.x = element_text(hjust = 0.4),
        legend.position="none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5)) +
  xlab("Publication year") +
  ylab("Proportion of articles in that journal") +
  lemon::coord_capped_cart(bottom='right')

@@

To compare with ecology we conducted a similar bibliometric analysis for archaeology journals indexed by the Web of Science service. We used the 'Cited Reference Search' to find articles citing R in the same way as above, and then refined the results to keep only those articles published during 2008-2018 that are included in the Web of Science category 'Archaeology'. This resulted in `r prettyNum(how_many_articles_in_top_archy_journals, ",")` articles, of which `r total_number_archaeology_articles_citing_r` cite R.Figure \@ref(fig:archaeology-journals-with-subplot) shows the temporal trend of citations of R in archaeological articles. There are three interesting details revealed by this figure.

```{r archaeology-journals-with-subplot, fig.cap=str_glue('Proportion of Archaeology articles per year citing R (a total of {total_number_archaeology_articles_citing_r} out of {prettyNum(how_many_articles_in_top_archy_journals, ",")} articles in our sample for 2008-2018). Labels to the right show the top five journals in our sample. Sub-plot shows articles published in the _Journal of Archaeological Science_ during 2008-2017. Data from Web of Science')}
# combine main plot and subplot
all_archaeology_citing_R_over_time_and_subplot <-
all_archaeology_citing_R_over_time  +
  annotation_custom(grob = ggplotGrob(jas_cites_r),
                          xmin = 2008, xmax = 2015,
                          ymin = 0.08, ymax = 0.20)
ggsave(plot = all_archaeology_citing_R_over_time_and_subplot,
       filename = "all_archaeology_citing_R_over_time_and_subplot.png",
       path = str_glue('{here::here()}/analysis/figures'),
       height = 9,
       width = 12)
knitr::include_graphics(str_glue('{here::here()}/analysis/figures/all_archaeology_citing_R_over_time_and_subplot.png'))
```

@@

As with data, preparing code to make it publicly available takes time to ensure that it is fit for others to read and use. Our casual observations of the ways in which archaeologists are using and sharing code show a high degree of variability in code style and organisation, indicating that currently most archaeologists are independently solving problems of how to write and share code through trial and error. We can use concepts from cultural evolution to understand this situation: learning by trial and error is known as guided variation in a cultural evolutionary framework. The dominance of guided variation suggests that using code is not yet widespread enough among archaeologists to propagate due to frequency-dependent biases. Frequency-dependent biases occur when people copy the most abundant variant, in this case the most commonly used tool for analysing data, in the population [cf. @boyd1988culture]. It also suggests that the professional benefits (i.e. the citation effects we describe above) are not yet widely known for code use to propagate due to content biases. Content biases result when some aspect of a variant's content, such as positive citation effects, makes it more likely to be adopted. Finally there may not yet enough highly-visible, prestigous researchers using code for this behaviour due to propagate due to model-based biases [@Rendell_et_al_2011; @boyd1988culture; @henrich2007dual]. This bias is based on imitation of highly prestigous, skilled or successful individuals. Considering the high variability of code using cultural evolutionary concepts, we conclude that there is unrealised potential to improve the efficiency of using and code in archaeology, firstly by converging on some widely-agreed upon conventions that will save researcher's time, and secondly by investing effort in communicating the professional benefits of using code.
